Sufficient Conditions for Reachability in Automata Networks with Priorities

### 2014-12-19 parts/example.tex
minor comments of the reviewers

\modMF{
This section aims at giving application examples of the static analysis method
that we developed in \pref{sec:sa}.
In \pref{ssec:ex-metazoan}, we apply our method to a small model of the
metazoan segmentation process,
and demonstrate how the flattening of \pref{ssec:flattening}
be used in such a case.
In \label{ssec:ex-tcrsig}, we apply our method to a large-scale model
of the T-cell receptor signalling pathway
in order to show the scalability of our method.
}

@@

    \path (1.8, 1.5) edge[superprio,coopupdate] (3.2, 2);
    \path (0.8, 4.5) edge[superprio,coopupdate] (3.2, 3);
  \end{tikzpicture}
  }
  \caption{
  \label{fig:metazoan-php}
    An example of PH $\oPH$ with $3$ classes of priorities,
    modelling the process of metazoan segmentation.
    Actions of $\ov{\PHh}^{(3)}$ (lowest priority) are in thin lines,
    actions of $\ov{\PHh}^{(2)}$ (intermediate priority) are in thick lines,
    and actions of $\ov{\PHh}^{(1)}$ (highest priority), whose only purpose is to update the cooperative sort $fc$,
    are not represented but modelled by the double dashed arrows.
  }
\end{figure}

@@

    \path (-0.8, 1.5) edge[prio,coopupdate] (-2.2, 1.5);
    \path (4.3, 4.5) edge[prio,coopupdate] (-2.2, 2.5);
  \end{tikzpicture}
  }
  \caption{
  \label{fig:metazoan-ph}
    An example of PH $\PH$ with $2$ classes of priorities,
    which is the flattening of the PH in \pref{fig:metazoan-php}.
    Actions of $\PHh^{(2)}$ (unprioritised) are in thin lines,
    and actions of $\PHh^{(1)}$ (prioritised) which are used to update cooperative sorts,
    are abstracted by thick dashed arrows.
  }
\end{figure}

@@

\modMF{
By analysing the detail of the results, one can find out that one of the output, NFkB, can never be activated.
In other words, FkB$_1$ is never reachable, whatever the state of the inputs,
while some combinations the other outputs (SRE, AP1, NFAT) can be activated for some configurations of inputs.
We thus want to check if these three outputs can be activated together,
that is, if the activation of one of the outputs does not prevent the activation of one of the others.
For this, as proposed in \pref{ssec:simult-ua},
we add a new cooperative sort $\total$ between SRE, AP1 and NFAT to the model,
and a sort $\reach$ with two processes ($\PHl_\reach = \{ \reach_0, \reach_1 \}$)
with the action $\PHhit{\total_{111}}{\reach_0}{\reach_1}$.
Finally, checking the reachability of $\reach_1$ in all configurations of the inputs
is always conclusive,
and the existence of positive answers allows to conclude that
it is possible to reach a state where SRE, AP1 and NFAT are simultaneously active.
}

>>>>

\modMF{
This section aims at giving application examples of the static analysis method
that we developed in \pref{sec:sa}.
In \pref{ssec:ex-metazoan}, we apply our method to a small model of the
metazoan segmentation process,
and demonstrate how the flattening of \pref{ssec:flattening}
be used in such a case.
In \pref{ssec:ex-tcrsig}, we apply our method to a large-scale model
of the T-cell receptor signalling pathway
in order to show the scalability of our method.
}

@@

    \path (1.8, 1.5) edge[superprio,coopupdate] (3.2, 2);
    \path (0.8, 4.5) edge[superprio,coopupdate] (3.2, 3);
  \end{tikzpicture}
  }
  \caption{
  \label{fig:metazoan-php}
    An example of PH $\oPH$ with $3$ classes of priorities,
    modelling the process of metazoan segmentation.
    Component $a$ models the pigment production, and is influenced by
    component $c$ that has the role of a clock,
    while $f$ represents the wavefront propagation.
    If component $a$ oscillates (that is, its active process changes regularly)
    then regular stripes are created on the metazoan.
    Actions of $\ov{\PHh}^{(3)}$ (lowest priority) are represented in thin lines,
    actions of $\ov{\PHh}^{(2)}$ (intermediate priority) are in thick lines,
    and actions of $\ov{\PHh}^{(1)}$ (highest priority), whose only purpose is to update the cooperative sort $fc$,
    are not explicitly represented but modelled by the double dashed arrows.
  }
\end{figure}

@@

    \path (-0.8, 1.5) edge[prio,coopupdate] (-2.2, 1.5);
    \path (4.3, 4.5) edge[prio,coopupdate] (-2.2, 2.5);
  \end{tikzpicture}
  }
  \caption{
  \label{fig:metazoan-ph}
    An example of PH $\PH$ with $2$ classes of priorities,
    which is the flattening of the PH in \pref{fig:metazoan-php}.
    In other words, this model has exactly the same dynamics as
    \pref{fig:metazoan-php}, but its actions make only two classes of priorities.
    Actions of $\PHh^{(2)}$ (unprioritised) are represented in thin lines,
    and actions of $\PHh^{(1)}$ (prioritised) which are used to update cooperative sorts,
    are abstracted by thick dashed arrows.
    The additional cooperative sort $fa$ is used to model the previous action
    $\PHhit{f_1}{c_0}{c_1}$,
    which belonged to the lowest class of priority and could be preempted 
    especially by action $\PHhit{fc_{10}}{a_0}{a_1}$.
  }
\end{figure}

@@

\modMF{
By analysing the detail of the results, one can find out that one of the outputs, NFkB, can never be activated.
In other words, FkB$_1$ is never reachable, whatever the state of the inputs,
while some combinations the other outputs (SRE, AP1, NFAT) can be activated for some configurations of inputs.
We thus want to check if these three outputs can be activated together,
that is, if the activation of one of the outputs does not prevent the activation of one of the others.
For this, as proposed in \pref{ssec:simult-ua},
we add a new cooperative sort $\total$ between SRE, AP1 and NFAT to the model,
and a sort $\reach$ with two processes ($\PHl_\reach = \{ \reach_0, \reach_1 \}$)
with the action $\PHhit{\total_{111}}{\reach_0}{\reach_1}$.
Finally, checking the reachability of $\reach_1$ in all configurations of the inputs
is always conclusive,
and the existence of positive answers allows to conclude that
it is possible to reach a state where SRE, AP1 and NFAT are simultaneously active.
}

### 2014-12-19 parts/intro.tex
forgot coloration in intro
minor comments of the reviewers

Discrete modelling frameworks for biological networks is an active research field where formal methods have proved that they were very powerful.
Such a work started in the seventies.
It was later enriched in many directions and widely used to elucidate many biological questions.
Among these questions, a major one is to understand precisely how biological systems evolve and behave; why and how they change their usual behaviours…
This leads to questions about the reachability (possible or inevitable) of some states.
The ultimate goal is to discover how it could be possible to prevent biological systems from reaching some pathological states.

Of course, such formal models on which analyses are performed are abstract representations of the actual studied systems.
They are associated with parameters that have to be synthesised %so as to be as much as possible fitting with the real systems having some observed behaviors.
to give the most faithful representation of the real systems with their observed behaviours.
As a matter of fact, the abstractions we get are more or less rough or accurate.
Usual formal frameworks for such modelling activities are state-transition systems or process algebras. % Petri nets
We developed a quite similar framework named the Process Hitting \cite{PMR10-TCSB},
consisting in a restriction where the evolution of a component is determined by the state of at most one other component that does not evolve.
In a sense, these kind of actions are of the form $X + Y \rightarrow X + Y'$ where $X$ behaves like a catalyst molecule that “hits” another molecule $Y$ and changes it into $Y'$, without being changed itself.
Assuming catalysts are always available, this can represent any biochemical system made of monomolecular reactions, and can also represent catalytic networks such as metabolic networks.
%, with the aim of avoiding to build the whole state space in order to be able to tackle very large systems
Our motivation behind this framework was to design a model and analysis techniques adapted to biological modelling.
These analyses avoid to build the whole state space, which allows to tackle very large systems (that would have led to a huge number of states, hopelessly too huge to be analysed).
They are based on the fact that most biological models have few levels of expression per component:
in Boolean networks~\cite{kauffman69,Thomas73} there are only two levels per component and in their multivalued equivalent, Asynchronous Discrete Networks~\cite{deJong02}, components rarely have more that four levels.

@@

\paragraph{Contributions}
Since our formalism (the Process Hitting) can be considered as a subset of Calculus of Communicating Systems, %is inspired from the $\pi$-calculus,
our work is related to such semantic ramifications of extending traditional process algebras with the concept of priority that allows for some transitions to be given precedence over others.
The concept is derived in two directions: dynamic versus static, the difference being naturally that the former one refers to a semantics where priority values may change during execution according to some evolution rules.
In our work, actions exhibit a static priority structure.
The actions modelling biological processes are divided in several classes of priorities which stand for the different reaction rates or relative preemptions.
Furthermore, the highest class of priority is used to model particular nonbiological actions, designated as “prioritised”,
that are used to encode Boolean functions inside the model.

This paper is an extension of \cite{FPMR13-CS2Bio} where we
introduced a new extension to the semantics of Process Hitting by partitioning actions into classes of priorities
(\pref{sec:ph}).
One of the objectives is to reach an accurate representation of cooperating components in the model, that was not fulfilled with the initial semantics, using only two classes of priorities.
In \pref{sec:sa},
we develop an efficient under-approximation of the reachability of the state of components on a subclass of this new framework, thus allowing to compute efficient static analysis.
This local reachability under-approximation can be also easily extended to study the reachability of a global state.
Furthermore, we provide a new extended under-approximation method to decide a successive reachability property.

@@

The main addition in this paper compared to \cite{FPMR13-CS2Bio} is
%two theorems, the first
\pref{thm:ordered-ua} that gives a method to cope with successive reachability,
and obtain more conclusive results in the case of several successive objectives.
Two other additions are 
\pref{th:bisimPHP} which states that any Process Hitting model with any number of classes of priorities
can be represented with only two classes of priorities,
thus extending the scope of our method to any possible Process Hitting model,
and \todo{def extraction}
which allows to extract a concrete scenario when our static analysis method
is conclusive.
%Both aim at extending the application range of the static analysis we developed.

\paragraph*{Notations}
%We denote: $\segm{a}{b} = \{ a, a+1, \dots, b-1, b \}$.

If $A$ is a finite set,
$|A|$ is the cardinality of $A$
and $\powerset(A)$ is the power set of $A$.
$\sN$ is the set of natural numbers,
$\sN^* = \sN \setminus \{ 0 \}$ is the set of positive natural numbers,
$\sNN = \sN \setminus \{ 0, 1 \}$ is the set of natural numbers strictly greater than 1,
and $\segm{x}{y} = \{ x, x+1, \dots, y-1, y \}$ is the set of natural numbers from $x$ to $y$ included.
If $x = (x_i)_{i \in \segm{1}{n}}$ is a sequence of elements indexed by $i \in \segm{1}{n}$,
%we denote %$|x| = (b-a)+1$ the size of this sequence and
$\indexes{x} = \segm{1}{n}$ is the set of indexes of this sequence.
We also denote by $\emptyseq$ the empty sequence.
If $A$ and $B$ are sets,
$f : A \rightarrow B$ denotes an application $f$ that maps the elements of $A$ to elements of $B$.
$\lfp{x_0}{x}{x'}$ is the least fixed point of the function $x \mapsto x'$ which is greater than $x_0$.
The Cartesian product is denoted $\times$.
If $z$ is a tuple of $n$ components, $\toset{z}$ denotes the corresponding set
	$\{z_1, \cdots, z_n\}$.

>>>>

Discrete modelling frameworks for biological networks is an active research field where formal methods have proved that they were very powerful.
Such a work started in the seventies.
It was later enriched in many directions and widely used to elucidate many biological questions.
Among these questions, a major one is to understand precisely how biological systems evolve and behave; why and how they change their usual behaviours, etc.
This leads to questions about the reachability (possible or inevitable) of some states.
The ultimate goal is to discover how it could be possible to prevent biological systems from reaching some pathological states.
\todo{Ajouter des références}

Of course, such formal models on which analyses are performed are abstract representations of the actual studied systems.
They are associated with parameters that have to be synthesised %so as to be as much as possible fitting with the real systems having some observed behaviors.
to give the most faithful representation of the real systems with their observed behaviours.
As a matter of fact, the abstractions we get are more or less rough or accurate.
Prevalent formal frameworks for such modelling activities are state-transition systems or process algebras. % Petri nets
We developed a quite similar framework named the Process Hitting \cite{PMR10-TCSB},
consisting in a restriction where the evolution of a component is determined by the state of at most one other component that does not evolve.
In a sense, these kind of actions are of the form $X + Y \rightarrow X + Y'$ where $X$ behaves like a catalyst molecule that “hits” another molecule $Y$ and changes it into $Y'$, without being changed itself.
Assuming catalysts are always available, this can represent any biochemical system made of monomolecular reactions, and can also represent catalytic networks such as metabolic networks.
%, with the aim of avoiding to build the whole state space in order to be able to tackle very large systems
Our motivation behind this framework was to design a model and analysis techniques adapted to biological modelling.
These analyses avoid to build the whole state space, which allows to tackle very large systems (that would have led to a huge number of states, hopelessly too huge to be analysed).
They are based on the fact that most biological models have few levels of expression per component:
in Boolean networks~\cite{kauffman69,Thomas73} there are only two levels per component and in their multivalued equivalent, Asynchronous Discrete Networks~\cite{deJong02}, components rarely have more than four levels.

@@

\subsection*{Contributions}
Since our formalism (the Process Hitting) can be considered as a subset of Calculus of Communicating Systems, %is inspired from the $\pi$-calculus,
our work is related to such semantic ramifications of extending traditional process algebras with the concept of priority that allows for some transitions to be given precedence over others.
The concept is derived in two directions: dynamic versus static, the difference being naturally that the former one refers to a semantics where priority values may change during execution according to some evolution rules.
In our work, actions exhibit a static priority structure.
The actions modelling biological processes are divided in several classes of priorities which stand for the different reaction rates or relative preemptions.
Furthermore, the highest class of priority is used to model particular nonbiological actions, designated as “prioritised”,
that are used to encode Boolean functions inside the model.

This paper is an extension of \cite{FPMR13-CS2Bio} where we
introduced a new extension to the semantics of Process Hitting by partitioning actions into classes of priorities
(\pref{sec:ph}).
One of the objectives is to reach an accurate representation of cooperating components in the model, that was not fulfilled with the initial semantics, using only two classes of priorities.
In \pref{sec:sa},
we develop an efficient under-approximation of the reachability of the state of components on a subclass of this new framework, thus allowing to compute efficient static analysis.
This local reachability under-approximation can be also easily extended to study the reachability of a global state.
Furthermore, we provide a new extended under-approximation method to check a successive reachability property.

@@

\modMF{
The main addition in this paper compared to \cite{FPMR13-CS2Bio} is
%two theorems, the first
\pref{thm:ordered-ua} that gives a method to cope with successive reachability,
and obtain more conclusive results in the case of several successive objectives.
Two other additions are 
\pref{th:bisimPHP} which states that any Process Hitting model with any number of classes of priorities
can be represented with only two classes of priorities,
thus extending the scope of our method to any possible Process Hitting model,
and \todo{def extraction}
which allows to extract a concrete scenario when our static analysis method
is conclusive.
%Both aim at extending the application range of the static analysis we developed.
}

\subsection*{Notations}
\label{notations}
%We denote: $\segm{a}{b} = \{ a, a+1, \dots, b-1, b \}$.

\paragraph*{Sets}
If $A$ is a finite set,
$|A|$ is the cardinality of $A$
and $\powerset(A)$ is the power set of $A$.
$\sN$ is the set of natural numbers,
$\sN^* = \sN \setminus \{ 0 \}$ is the set of positive natural numbers,
$\sNN = \sN \setminus \{ 0, 1 \}$ is the set of natural numbers strictly greater than 1,
and $\segm{x}{y} = \{ x, x+1, \dots, y-1, y \}$ is the set of natural numbers from $x$ to $y$ included.
The Cartesian product is denoted $\times$;
if $z$ is a tuple of $n$ components, $\toset{z}$ denotes the corresponding set:
$\toset{z} = \{z_1, \cdots, z_n\}$.

\paragraph*{Sequences}
If $n \in \sN$ and
$x = (x_i)_{i \in \segm{1}{n}}$ is a sequence of elements indexed by $i \in \segm{1}{n}$,
%we denote %$|x| = (b-a)+1$ the size of this sequence and
then $\indexes{x} = \segm{1}{n}$ is the set of indexes of this sequence.
Furthermore, if $a, b \in \indexes{x}$ with $a \leq b$,
then $x_{a..b} = (x_i)_{i \in \segm{a}{b}}$ is a subsequence of $x$.
%from element $n$ to element $m$ inclusive.
We also denote by $\emptyseq$ the empty sequence.

\paragraph*{Functions and least fixed point}
If $A$ and $B$ are sets,
$f : A \rightarrow B$ denotes an application $f$ that maps the elements of $A$ to elements of $B$.
Furthermore, if $f$ is a monotonically increasing and bounded function, then
$\lfp{x_0}{x}{x'}$ is the least fixed point of the function $x \mapsto x'$ which is greater than $x_0$.

### 2014-12-19 parts/models-ph-dn.tex
remove a lot of comments
TODO++ in the translation sections
minor comments of the reviewers

We exhibit an encoding of Asynchronous Discrete Networks (ADN) with the Process
Hitting using two classes of priorities
that respects all Conditions of \pref{ssec:hypothesis},
and prove a weak bisimulation relation.

@@

Let us denote the encoding of the ADN $\DNdef$ in Process Hitting with $2$ classes
of priorities by $\toPH\DNdef$ (\pref{def:DN2PH}).
For each component $i\in\segm{1}{n}$ of the ADN, two sorts are built: $a^i$ acting for the component
value, and $f^i$ acting for a cooperative sort between the components $\DNdep(f^i)$.
Sorts $a^i$ have one process $a^i_k$ per element in $k\in\mathbb F^i$.
Sorts $f^i$ have one process $f^i_\varsigma$ per state $\varsigma \in \underset{j\in\DNdep(f^i)}{\times}
\PHl_{a^j}$.
Two classes of actions are then defined:
$\Hits^{(1)}$ is the set of actions updating the cooperative sorts according to the current state of the
components:
if $j\in\DNdep(f^i)$, $a^j_k$ hits each process $f^i_\varsigma$ where $\get{\varsigma}{a^j}\neq
a^j_k$ to make it bounce to the process $f^i_{\varsigma \Cap a^j_k}$.
%$f^i_{\varsigma'}$ where $\get{\varsigma'}{a^j}=a^j_k$.
$\Hits^{(2)}$ is the set of actions encoding the transitions in the ADN:
$f^i_\varsigma$ hits the processes of sort $a^i$ to make them bounce to the process
$a^i_{k'}$ if and only if $k'=f^i(\decode \varsigma)$;
$\decode \varsigma$ being the ADN state corresponding to the PH (partial) state $\varsigma$ (note that
$f^i(\decode \varsigma)$ is fully defined because $\decode \varsigma$ specifies the state for all
the components in $\DNdep(f^i)$).

>>>>

We exhibit an encoding of Asynchronous Discrete Networks (ADN) with the Process
Hitting using two classes of priorities
and prove a weak bisimulation relation.
This translation is important as it allows to efficiently study the dynamics ADNs
by using the static analysis developed in \pref{sec:sa};
indeed, the bisimulation relation ensures that the dynamics of the
resulting PH model are accurately the same than the original.
Furthermore, the obtained model is ensured to respect
all Conditions of \pref{ssec:hypothesis}.

@@

Let us denote the encoding of the ADN $\DNdef$ in Process Hitting with $2$ classes
of priorities by $\toPH\DNdef$ (\pref{def:DN2PH}).
For each component $i\in\segm{1}{n}$ of the ADN, two sorts are built: $a^i$ acting for the component
value, and $f^i$ acting for a cooperative sort between the components $\DNdep(f^i)$.
Sorts $a^i$ have one process $a^i_k$ per element in $k\in\mathbb F^i$.
Sorts $f^i$ have one process $f^i_\varsigma$ per state $\varsigma \in \underset{j\in\DNdep(f^i)}{\times}
\PHl_{a^j}$.
Two classes of actions are then defined:
$\Hits^{(1)}$ is the set of actions updating the cooperative sorts according to the current state of the
components:
if $j\in\DNdep(f^i)$, $a^j_k$ hits each process $f^i_\varsigma$ where $\get{\varsigma}{a^j}\neq
a^j_k$ to make it bounce to the process $f^i_{\varsigma \Cap a^j_k}$.
$\Hits^{(2)}$ is the set of actions encoding the transitions in the ADN:
$f^i_\varsigma$ hits the processes of sort $a^i$ to make them bounce to the process
$a^i_{k'}$ if and only if $k'=f^i(\decode \varsigma)$;
$\decode \varsigma$ being the ADN state corresponding to the PH (partial) state $\varsigma$ (note that
$f^i(\decode \varsigma)$ is fully defined because $\decode \varsigma$ specifies the state for all
the components in $\DNdep(f^i)$).

### 2014-12-19 parts/models-ph-php.tex
TODO++ in the translation sections
minor comments of the reviewers

Consider a PH with $k \in \sNN$ classes of priories
$\ov{\PH} = (\ov{\PHs}; \ov{\PHl}; \ov{\PHa}^{\angles{k}})$
that respects all Conditions given in \pref{ssec:hypothesis}.
If $\oPH$ does not respect these Conditions,
it is then sufficient to instead consider the PH model with $k+1$ classes of priorities
$\ov{\PH}' = (\ov{\PHs}; \ov{\PHl}; \ov{\PHa}'^{\angles{k+1}})$
where $\ov{\PHh}'^{(1)} = \emptyset$.
The aim of this section is to propose a translation of $\oPH$
into a PH model with $2$ classes of priorities $\PH = (\PHs; \PHl; \PHa^{\angles{2}})$
called \emph{flattening},
which is weakly bisimilar
and that also respects all Conditions of \pref{ssec:hypothesis}.

>>>>

In this section, we give a translation that allows to obtain,
from any PH model with any number of priorities,
a PH model with 2 classes of priorities
that respects all Conditions of \pref{ssec:hypothesis}.
Such a translation is thus particularly useful to be able to study the dynamics of
any kind of PH model by using the static analysis developed in \pref{sec:sa},
which is not possible if these conditions are not met.

We consider in the following a PH with $k \in \sNN$ classes of priories
$\ov{\PH} = (\ov{\PHs}; \ov{\PHl}; \ov{\PHa}^{\angles{k}})$
that respects all Conditions given in \pref{ssec:hypothesis}.
If $\oPH$ does not respect these Conditions,
it is then sufficient to instead consider the PH model with $k+1$ classes of priorities
$\ov{\PH}' = (\ov{\PHs}; \ov{\PHl}; \ov{\PHa}'^{\angles{k+1}})$
where $\ov{\PHh}'^{(1)} = \emptyset$.
The aim of this section is to propose a translation of $\oPH$
into a PH model with $2$ classes of priorities $\PH = (\PHs; \PHl; \PHa^{\angles{2}})$
called \emph{flattening},
which is weakly bisimilar
and that also respects all Conditions of \pref{ssec:hypothesis}.
